\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Retrieval-augmented generation conditions retrieval on the user query, not on who the assistant is supposed to be. For role-specific assistants such as tutors, domain experts, and professional agents, this produces measurable identity drift across multi-turn dialogue. Two recent lines of work frame the intervention space: retrieval-side routing keyed on internal model states, and activation-side steering using extracted persona directions in the residual stream. The architectural gap between them, retrieve-on-drift gated by a persona-vector signal, motivated the v0.3 design of the Persona-RAG project.

This report presents two empirical experiments and an evaluation of what survives them. Experiment 1 replicates the persona-vector extraction methodology on Gemma-2-9B-Instruct at 4-bit NF4 quantisation across three hand-authored role personas and a four-layer middle band, with the two control conditions the Dubanowska defensive battery prescribes. Per-layer test \auroc{} reaches 1.000 on every (persona, layer) cell, with shuffled-label controls at chance and random-feature controls in the 0.58 to 0.65 band, well below the 0.70 weak floor. The persona-discriminative linear direction the published methodology extracts is real, and it replicates at quantised inference scale on a backbone not previously tested.

Experiment 2 tests whether that direction transfers to inference-time drift detection in multi-turn dialogue, the operational question every drift-gated retrieval mechanism depends on. The drift-trajectory sweep covers two backbones (Gemma-2-9B-Instruct, Llama-3.1-8B-Instruct) at 4-bit, two extraction regimes (prompt-scope and generation-scope), four layers per cell, three personas, and six hand-authored turns per condition. Across the 24 generation-scope cells and the 12 prompt-scope cells, the maximum absolute projection delta between in-persona and drifting conditions is 0.103, in the wrong direction, against a pre-registered proceed threshold of 0.30. The conclusion is that the contrast-trained linear direction does not transfer to the drift-detection question, even though the direction itself is real. This is the project's most substantive empirical finding.

What survives is M1, a typed-memory architecture with three separately-indexed stores (self-facts, worldview, episodic) and per-turn identity grounding (ID-RAG), and M3 in its v0.2 form, where the drift gate becomes an LLM-as-judge call rather than a persona-vector projection. We evaluate these against B1 (vanilla RAG) and B2 (a RoleGPT-level prompt baseline with dialogue few-shots) on a custom counterfactual-retrieval probe suite (90 multi-turn conversations across three personas and three probe types) scored by MiniCheck, SYCON, and a three-judge PoLL panel with human-validation calibration. B1 collapses under multi-turn pressure (PoLL persona-adherence 2.54 against the cluster's 4.5 to 4.7), B2 and M1 and M3 cluster within 0.20 points of each other on every probe type including the Type-B counterfactual-injection probes the mechanisms were designed to address, and M3's gate is precise but insensitive on this benchmark (precision 1.00, recall 0.06 at the calibrated threshold).

Three diagnostic ablations isolate the cluster's source: an oracle drift gate that fires on the known probe turns by construction, an M1-without-ID-RAG ablation that removes per-turn identity grounding, and a precision sweep from 4-bit NF4 to fp16 on the cs\_tutor persona. None of the three moves the headline cluster outside a pre-registered $\pm 0.15$ envelope. The binding constraint on persona-adherence quality at the 9B parameter scale is model capability, not retrieval architecture, gate calibration, or quantisation level.

The report's contribution is therefore threefold: a clean replication of persona-vector extraction at quantised inference scale on Gemma-2-9B (Experiment 1), a robust negative result on transfer to multi-turn drift detection (Experiment 2), and an architectural mechanism (M1 typed memory, M3 v0.2 LLM-judge gate) whose results map a measurable but capability-bound headroom above a strong prompt baseline. The negative result is the most actionable finding for subsequent work, because it removes one architectural assumption that prior literature did not test.
\bigskip

\noindent\textbf{Project source:} \url{https://github.com/yasinhessnawi1/Persona-RAG}
